\levelstay{Problem 2.4.4 (AO)}

Consider $n$ tagged particles of mass $m$ immersed in a fluid that is in thermal equilibrium. Let us assume that the concentration of these particles is vanishingly small, and furthermore, that they do not interact with one another. Under these assumptions, each particle moves independently of the others. Let $V = (v_1 + \ldots + v_n)/n$ be the velocity of their center of mass. \textbf{goal:} find the PDF of $V$ which we denote by $F^\text{eq}(V)$.

The PDF is given by
\begin{equation}
F^\text{eq}(V) = \int_{-\infty}^{\infty} dv_1 \cdots \int_{-\infty}^{\infty} dv_n ~ p^\text{eq}(v_1) \cdots p^\text{eq}(v_n) ~ \delta(V - \frac{1}{n}\sum_{i=1}^{n} v_i).
\end{equation}
Using the Fourier representation of the $\delta$-function
\begin{equation}
\delta(x) = \frac{1}{2\pi} \int_{-\infty}^{\infty} dp ~ \exp(i p x),
\end{equation}
we focus on the integral of 
\begin{equation}
\int_{-\infty}^{\infty} dv_1 \exp\bigg(-\frac{m v_1^2}{2 k T} -i p v_1/n\bigg) =  \bigg( \frac{2\pi k T}{m}\bigg)^{1/2} \times \exp\bigg(-\frac{p^2 k T }{2 n^2 m}\bigg).
\end{equation}
Notice that each of the $p^\text{eq}(v_j)$ terms comes with a normalization factor $\sqrt{m/2 \pi k T}$ that cancels with the first term in the expression above. This leaves us with one factor of $\exp(-p^2 k T/2 n^2 m)$ per particle. Repeating this process for all remaining particles and restoring the missing factor of $1/2\pi$ from the Fourier representation of the $\delta$-function yields the following integral:
\begin{eqnarray}
F^\text{eq}(V) &=& \frac{1}{2\pi}\int_{-\infty}^{\infty} dp ~ \exp\bigg(-\frac{p^2 k T }{2 n^2 m} \times n + i p V\bigg)\\
&=&  \boxed{\sqrt{\frac{m n}{2\pi k T}} ~ \exp\bigg(-\frac{V^2 m n }{2 k T} \bigg)}~.
\end{eqnarray}
This is means that the PDF of a linear combination of independent Gaussian random variables is again a Gaussian. Notice that the width of this distribution narrows with increasing $n$.